\documentclass[12pt]{article}

% --- Keep your existing libraries ---
\usepackage{geometry}
\geometry{
    a4paper,
    total={170mm,257mm},
    left=15mm,top=25mm
}

\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{hyperref}
\usepackage{theoremref}
\usepackage{fancyhdr}
% dhortcmds / dhortthms with portable fallbacks (the real packages win when present)
\IfFileExists{dhortcmds.sty}{\usepackage{dhortcmds}}{}
% portable fallbacks (no-ops when dhortcmds is present)

\IfFileExists{dhortthms.sty}{\usepackage{dhortthms}}{}
\usepackage{physics}
\usepackage[document]{ragged2e}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{centernot}
\usepackage{setspace}
\usepackage{array}
\usepackage{mdframed}
\usepackage{faktor}
\IfFileExists{bbm.sty}{\usepackage{bbm}}{}
\providecommand{\mathbbm}[1]{{\mathchoice{\mathrm{1}\mskip-4mu \mathrm{l}}{\mathrm{1}\mskip-4mu \mathrm{l}}{\mathrm{1}\mskip-4.5mu \mathrm{l}}{\mathrm{1}\mskip-5mu \mathrm{l}}}}
\usepackage{empheq}
\usepackage{tcolorbox}
\tcbuselibrary{most}
\newtcbox{\mathbox}{nobeforeafter, colback=white, colframe=black, boxrule=0.5pt, arc=3pt}

\newtcolorbox{mybox}{
    colback=white,
    colframe=black,
    boxrule=0.5pt,
    arc=4pt,
    left=6pt, right=6pt, top=4pt, bottom=4pt
}

% --- Readability tweaks ---
\AtBeginDocument{\fontsize{13.5}{17}\selectfont}
\setstretch{1.28}
\setlength{\parskip}{0.5em}
\setlength{\parindent}{0pt}
\allowdisplaybreaks

% --- Pagestyle Setup ---
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0.4pt}
\lhead{Classical Mechanics on Manifolds}
\rhead{\nouppercase{\leftmark}}
\cfoot{\thepage}

% Handy shorthand for the Lie derivative
\providecommand{\Lie}{\mathcal{L}}

\title{Geometric Formulation of Classical Mechanics}
\author{}
\date{}

\begin{document}
\maketitle
\thispagestyle{fancy}

\section{Introduction and Overview}\label{sec:headed}

Every student of physics encounters at least two formulations of classical mechanics. The first is codified in terms of Newton's laws of motion, where, for the most part, an individual's capability to solve a problem is capped by their capability to draw out all the forces involved, which in hindsight seems like the very antithesis of an exact science (it's inexact, dependent on getting the diagram right). In later years one is introduced to the power of Lagrangians and Hamiltonians, which render such doodling pointless and give a systematic way of deriving the equations of motion for a given system, allowing one to solve more complex problems, such as the double pendulum. However, the student is rarely told the mathematical basis beyond what a variation is with regard to these objects, i.e., they're seldom told what kind of mathematical object a Hamiltonian actually is. Where does it "live"? What is the deeper structure underlying such powerful machinery? The aim of these notes is to begin scratching the surface of these questions. It turns out that mechanics, at its heart, is geometric, and we shall soon see how and why.


An incentive to care about the deeper geometry of mechanics is that it helps us understand fundamental problems such as quantization better. In the last section, we will see that almost all the structures encountered in classical mechanics have a quantum analogue, which means we can hope to decipher more about quantum mechanics by finding analogies and similarities between the two.


A brief outline on what these notes will cover.
We start by discussing  the  \emph{configuration space} $\mathcal{Q}$ and its tangent bundle, where the Lagrangian lives (\S\ref{sec:config}). The Legendre transform (\S\ref{sec:legendre}) carries us to the Hamiltonian and to Hamilton's equations (\S\ref{sec:hamilton}), and in doing so relocates the theory to \emph{phase space}, the cotangent bundle $T^*\mathcal{Q}$ (\S\ref{sec:phasespace}). On phase space we meet the Poisson bracket and prove that bracket evolution and Hamiltonian evolution are the same thing (\S\ref{sec:poisson}). Then the geometry proper begins: after a minimal toolkit of differential forms (\S\ref{sec:forms}), we introduce the symplectic form (\S\ref{sec:symplectic}), rewrite Hamilton's equations without coordinates (\S\ref{sec:coordfree}), study flows and the vector fields that generate them (\S\S\ref{sec:flows}--\ref{sec:hamvf}), and generalize to abstract symplectic manifolds, where Darboux's theorem tells us something remarkable: symplectic geometry has no local invariants at all (\S\ref{sec:sympman}). Symplectic (canonical) transformations (\S\ref{sec:symptrans}) and Liouville's theorem (\S\ref{sec:liouville}) follow almost for free, and we close with statistical distributions on phase space and the Liouville equation (\S\ref{sec:statistical}), which is the precise classical twin of the von Neumann equation.

Conventions: configuration coordinates carry upper indices $q^a$, momenta lower indices $p_a$, with $a = 1, \dots, n$ for $n$ degrees of freedom. Repeated indices are summed (Einstein convention) unless said otherwise. Phase space coordinates are bundled as $\xi^\alpha$ with $\alpha = 1, \dots, 2n$. We work with time-independent Hamiltonians except where flagged.

\section{Configuration Space and the Tangent Bundle}\label{sec:config}

\subsection{Configuration space}

The \emph{configuration space} $\mathcal{Q}$ of a mechanical system is the set of all positions the system can occupy. For a single particle in ordinary space, the configuration space is isomorphic to $\mathcal{Q} = \mathbb{R}^3$; this seems unimpressive, the reader is probably familiar with solving problems in this space already. 

Things get interesting when we begin restricting degrees of freedom. Consider a planar pendulum whose position is completely specified by one angular coordinate $\theta$ with periodicity $2\pi$, i.e., $\theta$ and $\theta + 2\pi$ are the same physical position, so each position of this system can be mapped onto a point on the circle. The configuration space in this case is therefore a circle, i.e., $\mathcal{Q} = S^1$.





Similarly, a double pendulum has two angular coordinates, and in this case $\mc{Q} = S^1 \times S^1 = T^2$, a torus. To see why, notice that each angle $\theta_1$ and $\theta_2$ independently parametrizes a circle, since each pendulum arm returns to the same physical configuration under a shift of $2\pi$ therefore, the configuration space of the whole system is the set of all pairs $(\theta_1, \theta_2)$, which is  what the Cartesian product $S^1 \times S^1$ represents. In order to visualize better, imagine a circle (let this be all the values $\theta_1$ can take) and at each point of the circle attach another circle (all the values $\theta_2$ can take) this traces out a torus. 

It can also be shown that for a particle attached to a surface of a sphere has as its configuration space, $\mc{Q} = SO(3)$, which is the group of all rotation in 3-dimensional Eculedian space.


In each example above, we have taken the position data and found a neat way to arrange it, and in every example the resulting arrangement turned out to be a familiar 2D or 3D shape. This is not a coincidence. All these shapes have one thing in common: they are smooth $n$-dimensional \emph{manifolds}, spaces that locally look like $\mathbb{R}^n$ (in each case we chose an arbitrary origin and made up our own coordinates to help us keep track of the system; more on this when we discuss \emph{generalized coordinates}). This is important because we know how to do calculus on $\mathbb{R}^n$, and if a space, at close enough range, looks like $\mathbb{R}^n$, we might as well take it to be $\mathbb{R}^n$ and bring over the machinery of multivariable calculus. This area of mathematics is called \emph{differential geometry}. The $n$ in this context refers to the number of degrees of freedom a system has. Now that we have the position data neatly arranged in a way that lets us do calculus on it, it is natural to move on to higher-order quantities like velocity and acceleration in order to study the dynamics of the system.





\subsection{Velocities live in tangent spaces}

A configuration alone doesn't determine the future; we also need the velocity. Where does a velocity live? If the system traces a curve $q(t)$ through $\mathcal{Q}$, its velocity at time $t$ is the tangent vector $\dot{q}(t)$ to the curve. For the pendulum, at each point $\theta \in S^1$ the possible velocities $\dot\theta$ form a copy of $\mathbb{R}$: the tangent line to the circle at that point. Crucially, tangent vectors at \emph{different} points of a curved space can't be compared or added without further structure; a velocity at the bottom of the pendulum's swing and a velocity at the top live in different one-dimensional spaces.

So to each point $q \in \mathcal{Q}$ we attach the \emph{tangent space} $T_q\mathcal{Q}$, the $n$-dimensional vector space of all possible velocities at $q$. In coordinates, a tangent vector is specified by its components $\dot{q}^a$, and under a change of coordinates $q^a \to \tilde{q}^a(q)$ the components transform with the Jacobian,
\begin{flalign}
    \dot{\tilde{q}}^a = \pdv{\tilde{q}^a}{q^b}\,\dot{q}^b, \label{tangent-transform}
\end{flalign}
by the chain rule. (Upper index, Jacobian on the left: this is the defining transformation law of a \emph{contravariant} vector, and we'll shortly meet its mirror image.)

The collection of all tangent spaces, glued together over $\mathcal{Q}$,
\begin{flalign}
    T\mathcal{Q} \equiv \bigcup_{q \in \mathcal{Q}} \cbr{q} \times T_q\mathcal{Q},
\end{flalign}
is the \emph{tangent bundle}: a $2n$-dimensional manifold whose points are pairs (configuration, velocity). A point of $T\mathcal{Q}$ is exactly an allowed instantaneous state of the system in the Lagrangian sense. For the pendulum, $T\mathcal{Q} = S^1 \times \mathbb{R}$, an infinite cylinder: angle around, angular velocity along the axis. (The tangent bundle of the circle happens to be a global product; not every tangent bundle is, but locally they all are, and local is all we'll need.)

\subsection{The Lagrangian}

The Lagrangian is a function on the tangent bundle (possibly with explicit time dependence),
\begin{flalign}
    L : T\mathcal{Q} \times \mathbb{R} \to \mathbb{R}, \qquad (q, \dot{q}, t) \mapsto L(q, \dot{q}, t),
\end{flalign}
and the dynamics is selected by Hamilton's principle: physical trajectories between fixed endpoints extremize the action $S[q] = \int L\, \dd{t}$, which produces the Euler--Lagrange equations
\begin{flalign}
    \dv{t}\br{\pdv{L}{\dot{q}^a}} - \pdv{L}{q^a} = 0. \label{EL}
\end{flalign}
We take \eqref{EL} as known from the standard course and won't rederive it; our business here is what happens \emph{next}. For the mechanical systems we care most about, $L = T - V$ with $T$ quadratic in the velocities,
\begin{flalign}
    L(q,\dot{q}) = \frac{1}{2}\, g_{ab}(q)\, \dot{q}^a \dot{q}^b - V(q), \label{quadL}
\end{flalign}
where $g_{ab}(q)$ is a symmetric, positive-definite matrix of functions (a Riemannian metric on $\mathcal{Q}$, if we want to say it out loud: the kinetic energy defines a geometry on configuration space, which is a lovely story, but a different one from the geometry these notes are about).

\section{The Legendre Transform}\label{sec:legendre}

\subsection{The construction}

The passage from the Lagrangian to the Hamiltonian picture is the Legendre transform, and it deserves to be understood as a piece of mathematics in its own right rather than as the incantation ``$H = p\dot{q} - L$''. We follow Arnold's presentation, first in one variable.

Let $f : \mathbb{R} \to \mathbb{R}$ be convex, $f''(x) > 0$. The \emph{Legendre transform} of $f$ is a new function $g$ of a new variable $p$, built as follows. Fix $p$ and consider the line $y = px$. The function
\begin{flalign}
    F(p, x) = p\,x - f(x)
\end{flalign}
measures the vertical gap between the line and the graph of $f$. For fixed $p$, maximize over $x$: since $f$ is convex, $\pdv*{F}{x} = p - f'(x)$ has exactly one root $x(p)$, defined implicitly by
\begin{flalign}
    p = f'(x(p)), \label{leg-def}
\end{flalign}
and it's a maximum ($\pdv*[2]{F}{x} = -f''(x) < 0$). The Legendre transform is that maximal gap:
\begin{empheq}[box=\mathbox]{align}
    g(p) \equiv F(p, x(p)) = p\, x(p) - f(x(p)). \label{legendre}
\end{empheq}
Geometrically: $g(p)$ is (minus) the $y$-intercept of the tangent line to the graph of $f$ with slope $p$. The transform trades the description of a convex curve by its points $(x, f(x))$ for a description by its tangent lines $(p, g(p))$. Same curve, dual encoding.

Two quick examples to calibrate. For $f(x) = \frac{m x^2}{2}$: equation \eqref{leg-def} gives $p = mx$, so $x(p) = p/m$ and
\begin{flalign}
    g(p) = p\cdot\frac{p}{m} - \frac{m}{2}\frac{p^2}{m^2} = \frac{p^2}{2m}.
\end{flalign}
Kinetic energy in velocity form goes to kinetic energy in momentum form; this is the transform we'll be doing for a living. For $f(v) = e^v$: $p = e^v$ gives $v = \log p$ (for $p>0$) and $g(p) = p\log p - p$, which readers who've met entropy will recognize; the Legendre transform is also the engine room of thermodynamics, but again, different story.

\subsection{Involutivity: the transform undoes itself}

The single most important structural fact about the Legendre transform is that it's an \emph{involution}: applied twice, it gives back the original function. This is what guarantees that the Hamiltonian picture loses nothing, we can always Legendre-transform back and recover $L$. It deserves a proof, and Arnold's is short and slick.

\textbf{Theorem.} If $f$ is convex, then the Legendre transform $g$ of $f$ is convex, and the Legendre transform of $g$ is $f$.

\emph{Proof.} First, convexity of $g$. Differentiate \eqref{legendre} with respect to $p$, remembering that $x(p)$ moves:
\begin{flalign}
    g'(p) = x(p) + p\,x'(p) - f'(x(p))\,x'(p) = x(p) + \sbr{p - f'(x(p))}x'(p) = x(p), \label{leg-deriv}
\end{flalign}
where the bracket vanishes by the defining condition \eqref{leg-def}. (This little cancellation, the derivative through the implicit maximizer dropping out, is the envelope theorem, and it will happen to us again when we differentiate Hamiltonians.) So $g'(p) = x(p)$, and differentiating once more,
\begin{flalign}
    g''(p) = x'(p) = \dv{p}\sbr{(f')^{-1}(p)} = \frac{1}{f''(x(p))} > 0,
\end{flalign}
so $g$ is convex, and its Legendre transform is defined.

Now apply the transform to $g$. By definition, the transform of $g$ evaluated at the (new) variable $x$ is
\begin{flalign}
    G(x, p)\big|_{p = p(x)} \quad \text{with} \quad G(x,p) = x\,p - g(p), \quad x = g'(p(x)).
\end{flalign}
But we just computed $g'(p) = x(p)$: the condition $x = g'(p)$ is precisely the inversion of $p = f'(x)$. Substituting \eqref{legendre},
\begin{flalign}
    x\,p - g(p) = x\,p - \sbr{p\,x(p) - f(x(p))} = x\,p - p\,x + f(x) = f(x),
\end{flalign}
since at the stationary point $x(p) = x$. The transform of the transform is $f$. $\blacksquare$

There's an elegant symmetric way to say all this. For any $x$ and any $p$ (not necessarily matched), $g(p) \geq px - f(x)$ by the maximality in \eqref{legendre}, i.e.
\begin{flalign}
    p\,x \;\leq\; f(x) + g(p), \label{young}
\end{flalign}
with equality precisely when $p = f'(x)$. This is \emph{Young's inequality}, and it exhibits $f$ and $g$ as perfectly dual: each is the best possible function making \eqref{young} tight. (Try $f = x^\alpha/\alpha$, $g = p^\beta/\beta$ with $\frac{1}{\alpha} + \frac{1}{\beta} = 1$: the classical Young inequality of analysis drops out.)

\subsection{Many variables, and the physical case}

For a convex function of several variables (positive-definite Hessian), everything goes through verbatim with $x \to (x^1,\dots,x^n)$ and the transform performed in all variables at once:
\begin{flalign}
    p_a = \pdv{f}{x^a}, \qquad g(p) = p_a x^a(p) - f(x(p)).
\end{flalign}
The transform can also be taken in \emph{some} variables while the others sit as spectators; that's exactly our situation. Given a Lagrangian $L(q, \dot{q}, t)$, convex in the velocities (which \eqref{quadL} is, since $g_{ab}$ is positive definite), define the \emph{canonical momenta}
\begin{empheq}[box=\mathbox]{align}
    p_a \equiv \pdv{L}{\dot{q}^a}, \label{canmom}
\end{empheq}
invert to get $\dot{q}^a(q, p, t)$, and set
\begin{empheq}[box=\mathbox]{align}
    H(q, p, t) \equiv p_a\, \dot{q}^a(q,p,t) - L\br{q, \dot{q}(q,p,t), t}. \label{hamdef}
\end{empheq}
The Hamiltonian is the Legendre transform of the Lagrangian in the velocity slots, with $q$ and $t$ along for the ride. Involutivity guarantees the two descriptions carry identical information.

Notice, and this matters more than it looks, that the momentum \eqref{canmom} carries a \emph{lower} index. Under a change of configuration coordinates, the chain rule gives
\begin{flalign}
    \tilde{p}_a = \pdv{L}{\dot{\tilde{q}}^a} = \pdv{L}{\dot{q}^b}\pdv{\dot{q}^b}{\dot{\tilde{q}}^a} = p_b\, \pdv{q^b}{\tilde{q}^a}, \label{cotangent-transform}
\end{flalign}
the \emph{inverse} Jacobian compared to \eqref{tangent-transform}. Momenta are not velocities in disguise; they transform oppositely. They are \emph{covectors}, and we'll take that observation seriously in \S\ref{sec:phasespace}.

For the quadratic Lagrangian \eqref{quadL}, everything is explicit: $p_a = g_{ab}\dot{q}^b$, so $\dot{q}^a = g^{ab}p_b$ with $g^{ab}$ the inverse matrix, and
\begin{flalign}
    H = p_a g^{ab} p_b - \frac{1}{2}g_{cd}\, g^{ca}p_a\, g^{db}p_b - \br{-V} = \frac{1}{2}\, g^{ab}\, p_a p_b + V(q), \label{quadH}
\end{flalign}
using $g_{cd}g^{ca}g^{db} = \delta_d^a g^{db} = g^{ab}$. The Hamiltonian is the total energy $T + V$, with the kinetic term now built from the \emph{inverse} metric. (We'll redo this calculation as a worked exercise from the Hamilton's-equations side in \S\ref{sec:hamilton}.)

\section{Hamilton's Equations}\label{sec:hamilton}

\subsection{The derivation: compute $\dd{H}$ two ways}

Hamilton's equations fall out of a single clean move: compute the differential of $H$ twice, once from the definition \eqref{hamdef} and once from the fact that $H$ is a function of $(q, p, t)$, then compare coefficients.

From the definition, treating all of $q, \dot{q}, p, t$ as varying (the constraint that $\dot q$ is the function $\dot q(q,p,t)$ will take care of itself),
\begin{flalign}
    \dd{H} &= \dd{\br{p_a \dot{q}^a - L}} \notag \\
    &= \dot{q}^a\, \dd{p_a} + p_a\, \dd{\dot{q}^a} - \pdv{L}{q^a}\dd{q^a} - \pdv{L}{\dot{q}^a}\dd{\dot{q}^a} - \pdv{L}{t}\dd{t} \notag \\
    &= \dot{q}^a\, \dd{p_a} + \underbrace{\br{p_a - \pdv{L}{\dot{q}^a}}}_{=\;0 \text{ by } \eqref{canmom}}\dd{\dot{q}^a} - \pdv{L}{q^a}\dd{q^a} - \pdv{L}{t}\dd{t} \notag \\
    &= \dot{q}^a\, \dd{p_a} - \pdv{L}{q^a}\,\dd{q^a} - \pdv{L}{t}\,\dd{t}. \label{dH-way1}
\end{flalign}
The envelope-theorem cancellation strikes again: the $\dd{\dot{q}^a}$ terms drop out exactly because $p$ was \emph{defined} as $\pdv*{L}{\dot{q}}$. This is why the Legendre transform is the right change of variables and not just \emph{a} change of variables: it's the unique one for which the velocity differentials disappear, leaving $H$ a natural function of $(q,p,t)$ alone.

On the other hand, since $H = H(q, p, t)$,
\begin{flalign}
    \dd{H} = \pdv{H}{q^a}\dd{q^a} + \pdv{H}{p_a}\dd{p_a} + \pdv{H}{t}\dd{t}. \label{dH-way2}
\end{flalign}
The differentials $\dd{q^a}, \dd{p_a}, \dd{t}$ are independent, so \eqref{dH-way1} and \eqref{dH-way2} match term by term:
\begin{flalign}
    \pdv{H}{p_a} = \dot{q}^a, \qquad \pdv{H}{q^a} = -\pdv{L}{q^a}, \qquad \pdv{H}{t} = -\pdv{L}{t}. \label{match}
\end{flalign}
So far no equations of motion have been used; \eqref{match} is pure Legendre bookkeeping. Now feed in the dynamics. On a physical trajectory the Euler--Lagrange equation \eqref{EL} says $\pdv*{L}{q^a} = \dv*{p_a}{t} = \dot{p}_a$, and the middle relation of \eqref{match} becomes $\pdv*{H}{q^a} = -\dot{p}_a$. Together with the first relation:
\begin{empheq}[box=\mathbox]{align}
    \dot{q}^a = \pdv{H}{p_a}, \qquad \dot{p}_a = -\pdv{H}{q^a}. \label{hameqs}
\end{empheq}
These are \emph{Hamilton's equations}: $2n$ first-order ODEs replacing the $n$ second-order Euler--Lagrange equations. The trade is even. What's been gained is structure, and the whole remainder of these notes is about saying precisely what that structure is.

A first taste of the structure: the equations are almost symmetric between $q$ and $p$, spoiled only by a single minus sign. Hold that thought; that minus sign \emph{is} the symplectic form (\S\ref{sec:symplectic}).

\subsection{Energy conservation}

Take the total time derivative of $H$ along a trajectory:
\begin{flalign}
    \dv{H}{t} = \pdv{H}{q^a}\dot{q}^a + \pdv{H}{p_a}\dot{p}_a + \pdv{H}{t}
    = \pdv{H}{q^a}\pdv{H}{p_a} - \pdv{H}{p_a}\pdv{H}{q^a} + \pdv{H}{t} = \pdv{H}{t}, \label{Hdot}
\end{flalign}
substituting \eqref{hameqs} in the middle step. The first two terms cancel \emph{identically}, by the antisymmetric structure of Hamilton's equations, not by any property of the particular $H$. So if $H$ has no explicit time dependence ($\pdv*{L}{t} = 0$, hence $\pdv*{H}{t} = 0$ by \eqref{match}), then $H$ is constant along every trajectory: energy conservation. We'll re-derive this in one line, twice more, from progressively more geometric standpoints (\S\ref{sec:poisson}, \S\ref{sec:coordfree}); watching the same fact become more and more obvious is a good way to measure what the geometry buys.

\subsection{Worked exercise: the quadratic Hamiltonian}

Let's run Hamilton's equations on \eqref{quadH}, $H = \frac{1}{2}g^{ab}(q)\,p_a p_b + V(q)$, and check we recover the Euler--Lagrange dynamics of \eqref{quadL}. First equation:
\begin{flalign}
    \dot{q}^a = \pdv{H}{p_a} = \frac{1}{2}g^{cb}\br{\delta^a_c\, p_b + p_c\, \delta^a_b} = g^{ab}p_b,
\end{flalign}
using the symmetry of $g^{ab}$; this correctly inverts $p_a = g_{ab}\dot{q}^b$. Second equation:
\begin{flalign}
    \dot{p}_a = -\pdv{H}{q^a} = -\frac{1}{2}\pdv{g^{bc}}{q^a}\,p_b p_c - \pdv{V}{q^a}.
\end{flalign}
To compare with Euler--Lagrange, differentiate $p_a = g_{ab}\dot{q}^b$ in time and use $\pdv*{L}{q^a} = \frac{1}{2}\pdv*{g_{bc}}{q^a}\,\dot{q}^b\dot{q}^c - \pdv*{V}{q^a}$; matching the two requires the identity $\pdv*{g^{bc}}{q^a} = -g^{bd}g^{ce}\,\pdv*{g_{de}}{q^a}$ (differentiate $g^{bd}g_{dc} = \delta^b_c$), and then the two forms agree line by line. It's a worthwhile ten minutes of index gymnastics; the point of recording it is that the Hamiltonian formalism handles position-dependent kinetic terms (curvilinear coordinates, constrained systems) with no special pleading.

\begin{mybox}
\textbf{A caveat about ``$H = $ energy''.} The Hamiltonian \eqref{hamdef} is the Legendre transform of $L$; it equals the mechanical energy $T + V$ when $L = T - V$ with $T$ purely quadratic in velocities. Both hypotheses can fail. The charged particle in a magnetic field has $L = \frac{m}{2}\dot{\vb{x}}^2 + \frac{e}{c}\dot{\vb{x}}\cdot\vb{A}(\vb{x}) - e\phi(\vb{x})$, linear-plus-quadratic in velocity; the canonical momentum $\vb{p} = m\dot{\vb{x}} + \frac{e}{c}\vb{A}$ is \emph{not} $m\dot{\vb{x}}$, and $H = \frac{1}{2m}\br{\vb{p} - \frac{e}{c}\vb{A}}^2 + e\phi$ is the energy only after this substitution. The lesson generalizes: the canonical momentum is defined by \eqref{canmom} and by nothing else, and confusing it with $m\dot{q}$ is the classic error this formalism punishes.
\end{mybox}

\section{Phase Space: The Cotangent Bundle}\label{sec:phasespace}

\subsection{Momenta are covectors}

Hamilton's equations live on the space of pairs $(q, p)$, and it's time to say what that space is, geometrically. We noticed in \eqref{cotangent-transform} that momenta transform with the inverse Jacobian: $\tilde{p}_a = p_b\, \pdv*{q^b}{\tilde{q}^a}$. Objects with that transformation law have a name. A \emph{covector} (or one-form) at a point $q \in \mathcal{Q}$ is a linear map $T_q\mathcal{Q} \to \mathbb{R}$: a machine that eats a velocity and returns a number. The pairing
\begin{flalign}
    p(\dot{q}) = p_a\, \dot{q}^a
\end{flalign}
is coordinate-independent precisely because the Jacobians of \eqref{tangent-transform} and \eqref{cotangent-transform} cancel: $\tilde{p}_a \dot{\tilde{q}}^a = p_b \pdv*{q^b}{\tilde{q}^a}\, \pdv*{\tilde{q}^a}{q^c}\,\dot{q}^c = p_b\, \delta^b_c\, \dot{q}^c = p_a \dot{q}^a$. And physics hands us this pairing directly: $p_a \dot{q}^a$ is the first term of the Hamiltonian \eqref{hamdef}, and $p_a\,\dd{q^a}$ will shortly be the object whose derivative is the symplectic form. Momentum's job in life is to be paired with velocity; it is a covector by profession, not by accident.

The space of covectors at $q$ is the \emph{cotangent space} $T^*_q\mathcal{Q}$ (the dual vector space to $T_q\mathcal{Q}$), and the bundle of all of them,
\begin{empheq}[box=\mathbox]{align}
    \Gamma \equiv T^*\mathcal{Q} = \bigcup_{q\in\mathcal{Q}} \cbr{q}\times T^*_q\mathcal{Q},
\end{empheq}
is the \emph{cotangent bundle}: phase space. It is a $2n$-dimensional manifold, coordinatized locally by $(q^1,\dots,q^n,p_1,\dots,p_n)$, and a single point of it is a complete instantaneous state of the system: through each point of phase space passes exactly one solution of Hamilton's equations (existence and uniqueness for first-order ODEs). That's the operational meaning of ``state'' and the reason phase space, not configuration space, is the true home of mechanics.

For the pendulum: $\Gamma = S^1 \times \mathbb{R}$, the cylinder again, but now angle and \emph{angular momentum}. The distinction with $T\mathcal{Q}$ looks pedantic here (both are cylinders), and indeed whenever a metric is around one can convert velocities to momenta with $g_{ab}$; but the two bundles carry different natural structures. The cotangent bundle has one the tangent bundle lacks: a canonical symplectic form, present before we choose any metric, Lagrangian, or Hamiltonian. Getting to it is the goal of \S\ref{sec:symplectic}.

\subsection{Unified coordinates and the symplectic matrix}

For everything that follows it pays enormously to stop segregating $q$'s from $p$'s and bundle them into single phase space coordinates
\begin{flalign}
    \xi^\alpha = \br{q^1, \dots, q^n,\; p_1, \dots, p_n}, \qquad \alpha = 1, \dots, 2n. \label{xidef}
\end{flalign}
(An index at the top for all of them; from the phase-space point of view the $\xi^\alpha$ are just coordinates, and the $q$-vs-$p$ distinction is bookkeeping we're about to encode in a matrix instead.) Hamilton's equations \eqref{hameqs} read, in the two halves,
\begin{flalign}
    \dot{\xi}^a = \pdv{H}{\xi^{n+a}}, \qquad \dot{\xi}^{n+a} = -\pdv{H}{\xi^{a}}, \qquad a = 1,\dots,n,
\end{flalign}
which assemble into the single equation
\begin{empheq}[box=\mathbox]{align}
    \dot{\xi}^\alpha = \tilde{\Omega}^{\alpha\beta}\, \pdv{H}{\xi^\beta}, \qquad
    \tilde{\Omega} = \begin{pmatrix} 0 & \mathbbm{1}_n \\ -\mathbbm{1}_n & 0 \end{pmatrix}. \label{hameqs-xi}
\end{empheq}
The constant antisymmetric matrix $\tilde{\Omega}$ is the \emph{symplectic matrix}. All of Hamiltonian mechanics' asymmetry between position and momentum, the lone minus sign we flagged under \eqref{hameqs}, now lives in this one matrix, and the equations themselves are manifestly uniform in the $2n$ variables. Two algebraic properties we'll use constantly:
\begin{flalign}
    \tilde{\Omega}^{\mathrm{T}} = -\tilde{\Omega}, \qquad \tilde{\Omega}^2 = -\mathbbm{1}_{2n}, \label{omega-props}
\end{flalign}
the second by block multiplication. So $\tilde{\Omega}$ is antisymmetric and squares to minus one: it is, among other things, a linear complex structure on $\mathbb{R}^{2n}$, which is a first hint of the deep kinship between symplectic geometry and complex geometry (a hint we'll leave as a hint).

Equation \eqref{hameqs-xi} is worth staring at. It says: the phase-space velocity field of the dynamics is \emph{the gradient of $H$, rotated by $\tilde{\Omega}$}. In ordinary gradient flow, $\dot{\xi} = -\nabla H$, trajectories run downhill and $H$ decreases. Here the ``gradient'' is turned sideways before use, so trajectories run along level sets of $H$ instead of across them, and $H$ is conserved. One matrix, and dissipation becomes conservation. This picture, dynamics as symplectically-rotated gradient flow, is the coordinate avatar of the coordinate-free statement $i_X\Omega = -\dd{H}$ toward which we're building.

\section{Poisson Brackets}\label{sec:poisson}

\subsection{Definition and first properties}

Observables of classical mechanics are smooth functions on phase space, $f : \Gamma \to \mathbb{R}$ (energy, angular momentum, the coordinates themselves). How does an observable change along the motion? Chain rule plus \eqref{hameqs-xi}:
\begin{flalign}
    \dv{f}{t} = \pdv{f}{\xi^\alpha}\,\dot{\xi}^\alpha = \pdv{f}{\xi^\alpha}\,\tilde{\Omega}^{\alpha\beta}\,\pdv{H}{\xi^\beta}.
\end{flalign}
The bilinear combination on the right is the star of the section. Define, for any two observables $f, g$, the \emph{Poisson bracket}
\begin{empheq}[box=\mathbox]{align}
    \cbr{f, g} \equiv \pdv{f}{\xi^\alpha}\,\tilde{\Omega}^{\alpha\beta}\,\pdv{g}{\xi^\beta}
    = \sum_{a=1}^n\br{\pdv{f}{q^a}\pdv{g}{p_a} - \pdv{f}{p_a}\pdv{g}{q^a}}, \label{PB}
\end{empheq}
where the second form just unpacks the blocks of $\tilde{\Omega}$. Then the time evolution of any observable, along any Hamiltonian flow, is
\begin{empheq}[box=\mathbox]{align}
    \dv{f}{t} = \cbr{f, H} + \pdv{f}{t}, \label{PB-evolution}
\end{empheq}
where the last term is present only if $f$ depends on time explicitly. Two immediate dividends. Setting $f = \xi^\alpha$ recovers Hamilton's equations themselves, $\dot{\xi}^\alpha = \cbr{\xi^\alpha, H}$, so \eqref{PB-evolution} contains all of the dynamics. And setting $f = H$ (time-independent) gives $\dot{H} = \cbr{H, H} = 0$ by antisymmetry: energy conservation, second derivation, now one line.

On the fundamental coordinates the bracket evaluates to
\begin{flalign}
    \cbr{\xi^\alpha, \xi^\beta} = \delta^\alpha{}_\mu\, \tilde{\Omega}^{\mu\nu}\, \delta^\beta{}_\nu = \tilde{\Omega}^{\alpha\beta},
    \qquad\text{i.e.}\qquad
    \cbr{q^a, p_b} = \delta^a_b, \quad \cbr{q^a, q^b} = \cbr{p_a, p_b} = 0, \label{fundamental-PB}
\end{flalign}
the \emph{fundamental Poisson brackets}, whose quantum echo $[\hat{q}^a, \hat{p}_b] = i\hbar\,\delta^a_b$ needs no introduction.

Now the structural properties. For all observables $f, g, h$ and constants $\lambda$:
\begin{enumerate}[label=(\roman*)]
    \item \textbf{Antisymmetry:} $\cbr{f, g} = -\cbr{g, f}$.
    \item \textbf{Bilinearity:} $\cbr{f + \lambda g, h} = \cbr{f,h} + \lambda\cbr{g,h}$, and similarly in the second slot.
    \item \textbf{Leibniz rule:} $\cbr{fg, h} = f\cbr{g,h} + \cbr{f,h}g$.
    \item \textbf{Jacobi identity:} $\cbr{f,\cbr{g,h}} + \cbr{g,\cbr{h,f}} + \cbr{h,\cbr{f,g}} = 0$.
\end{enumerate}
Antisymmetry is the antisymmetry of $\tilde{\Omega}$ read through \eqref{PB}:
\begin{flalign}
    \cbr{g,f} = \partial_\alpha g\, \tilde{\Omega}^{\alpha\beta} \partial_\beta f = \partial_\beta f\, \tilde{\Omega}^{\alpha\beta} \partial_\alpha g = -\partial_\beta f\, \tilde{\Omega}^{\beta\alpha}\partial_\alpha g = -\cbr{f,g}
\end{flalign}
(relabel and flip). Bilinearity is the linearity of derivatives. Leibniz is the product rule: $\partial_\alpha(fg) = f\partial_\alpha g + (\partial_\alpha f) g$, inserted into \eqref{PB}. These three say that $\cbr{\,\cdot\,, h}$ is, for each fixed $h$, a \emph{derivation}: it acts on the algebra of observables the way a first-order differential operator does. That's not a metaphor; we'll construct the differential operator explicitly in \S\ref{sec:hamvf} (it's the Hamiltonian vector field of $h$).

Jacobi is the only one that requires actual work, and brute-force expansion is a famously bloody affair. Here is Arnold's argument, which gets it with almost no computation. Consider the first term, $\cbr{f, \cbr{g,h}}$: it's a first-order differential operator applied to $\cbr{g,h}$, which is itself built from first derivatives of $g$ and $h$. So expanded out, $\cbr{f,\cbr{g,h}}$ is a sum of terms each containing \emph{second} derivatives of $g$ or of $h$ (times first derivatives of the others), and no second derivatives of $f$. Now look at the full cyclic sum and collect all terms containing second derivatives of $h$. They can come only from the first two terms of the cycle,
\begin{flalign}
    \cbr{f, \cbr{g, h}} + \cbr{g, \cbr{h, f}} = \cbr{f,\cbr{g,h}} - \cbr{g, \cbr{f, h}},
\end{flalign}
and this combination is the \emph{commutator} of the two first-order operators $\cbr{f, \,\cdot\,}$ and $\cbr{g, \,\cdot\,}$ acting on $h$. But the commutator of two first-order differential operators is again first-order: writing $A = A^\alpha \partial_\alpha$ and $B = B^\beta\partial_\beta$,
\begin{flalign}
    [A, B] = A^\alpha \partial_\alpha (B^\beta \partial_\beta) - B^\beta\partial_\beta(A^\alpha \partial_\alpha)
    = \br{A^\alpha\partial_\alpha B^\beta - B^\alpha \partial_\alpha A^\beta}\partial_\beta,
\end{flalign}
because the would-be second-derivative pieces $A^\alpha B^\beta \partial_\alpha\partial_\beta$ appear in both products with the same sign and cancel in the difference. So the cyclic sum contains no second derivatives of $h$. By the manifest cyclic symmetry of the expression, it likewise contains no second derivatives of $f$ or of $g$. But every term of the expansion contains a second derivative of something. A sum of terms, each of which must vanish by this counting, is zero. $\blacksquare$

(File the mechanism away: ``commutators of derivations are derivations'' is doing the heavy lifting, and it is exactly the same mechanism that makes commutators of quantum observables close among themselves.)

\begin{mybox}
\textbf{Sign conventions, a genuine hazard.} We follow the convention \eqref{PB}, in which $\cbr{q, p} = +1$ and evolution reads $\dot{f} = \cbr{f, H}$. \emph{Arnold uses the opposite sign}: his bracket is $(f, g) = $ derivative of $f$ along the flow of $g$, which works out to $(f,g) = -\cbr{f,g}$ in our convention; his evolution law and his relation between brackets and commutators of flows carry compensating signs, and everything is internally consistent, but transplanting an equation from Arnold without checking the sign is a reliable way to lose an afternoon. The same alertness will be needed at the quantum interface: Dirac's quantum Poisson bracket is the commutator \emph{divided by} $i\hbar$, so $\cbr{f,g} \to \frac{1}{i\hbar}[\hat{f},\hat{g}]$, and the $i\hbar$ is part of the dictionary, not optional.
\end{mybox}

\subsection{The converse: bracket evolution is exactly Hamiltonian evolution}

Equation \eqref{PB-evolution} says every Hamiltonian flow is a bracket flow. It's worth asking the converse, because the answer is a genuine theorem and it delimits precisely what the bracket formulation contains. Suppose someone hands us an evolution law of bracket type,
\begin{flalign}
    \dv{f}{t} = \cbr{f, G} \quad \text{for all observables } f, \label{bracket-evo}
\end{flalign}
for some phase-space function $G$: then it's Hamiltonian by construction, with $H = G$. The sharper question: suppose we're handed a \emph{flow}, a velocity field $\dot{\xi}^\alpha = X^\alpha(\xi)$ on phase space. When is it of the form \eqref{bracket-evo} for some $G$?

Taking $f = \xi^\alpha$ in \eqref{bracket-evo} gives $X^\alpha = \tilde{\Omega}^{\alpha\beta}\partial_\beta G$, or, multiplying by $\tilde{\Omega}$ and using $\tilde{\Omega}^2 = -\mathbbm{1}$ from \eqref{omega-props},
\begin{flalign}
    \partial_\alpha G = -\tilde{\Omega}_{\alpha\beta}\, X^\beta \equiv \lambda_\alpha, \label{lambda-def}
\end{flalign}
where we write $\tilde{\Omega}_{\alpha\beta}$ for the (numerically identical, in these coordinates) inverse matrix. So the question is: when is the covector field $\lambda_\alpha$ built from $X$ a gradient? Locally, the answer is the classical integrability condition: $\lambda_\alpha = \partial_\alpha G$ has a solution in a neighbourhood if and only if
\begin{flalign}
    \partial_\alpha \lambda_\beta - \partial_\beta \lambda_\alpha = 0, \label{integrability}
\end{flalign}
i.e. the antisymmetrized derivative vanishes (necessity is the symmetry of second partials, $\partial_\alpha\partial_\beta G = \partial_\beta\partial_\alpha G$; sufficiency is the Poincar\'e lemma, and the explicit solution is the line-integral construction $G(\xi) = \int_0^1 \dd{s}\, \lambda_\alpha(s\xi)\,\xi^\alpha$ in a star-shaped neighbourhood, exactly the construction familiar from ``when is a force field a potential field''). Condition \eqref{integrability}, unpacked via \eqref{lambda-def}, is a first-order differential condition on the velocity field $X$: it says $X$ must preserve the symplectic structure, in a sense we'll make exact in \S\ref{sec:hamvf} (where \eqref{integrability} will reappear as $\dd{(i_X\Omega)} = 0$).

So: \emph{a flow is a bracket flow if and only if it's locally a Hamiltonian flow}, and the function $G$ generating it is unique up to an additive constant (two solutions of \eqref{lambda-def} differ by a function with vanishing gradient). The word ``locally'' is not decorative: on a topologically nontrivial phase space, condition \eqref{integrability} can hold globally while no single global $G$ exists (the line integral defining $G$ can pick up different values along non-contractible loops). Such flows are called \emph{locally Hamiltonian}, and the gap between locally and globally Hamiltonian is measured by the topology of $\Gamma$, its first de Rham cohomology, to give the gap its proper name. For $\Gamma = \mathbb{R}^{2n}$ the gap is empty and the two notions coincide.

\subsection{Poisson's theorem}

A \emph{constant of motion} is an observable with $\cbr{f, H} = 0$ (and no explicit time dependence). The Jacobi identity has an immediate and useful consequence:

\textbf{Theorem (Poisson).} If $f$ and $g$ are constants of motion, so is $\cbr{f, g}$.

\emph{Proof.} By Jacobi, $\cbr{\cbr{f,g}, H} = -\cbr{\cbr{g,H},f} - \cbr{\cbr{H,f},g} = -\cbr{0, f} - \cbr{0, g} = 0$. $\blacksquare$

Occasionally this manufactures genuinely new conservation laws (two components of angular momentum bracket to the third); more often it returns zero or a function of the ones already in hand. Either way, structurally it says the constants of motion form a \emph{closed subalgebra} under the bracket, which is the seed of the entire integrability story (Liouville--Arnold), a story for another set of notes.


\section{A Minimal Kit of Differential Forms}\label{sec:forms}

Before the symplectic form can be introduced honestly we need a small amount of machinery: differential forms and the operations on them. This section is the toolkit, assembled quickly but with the definitions stated in full; readers fluent in forms should skim for notation and move on.

\subsection{Forms}

A \emph{differential $k$-form} on a manifold $M$ is a smooth field of antisymmetric multilinear maps: at each point, $\omega$ eats $k$ tangent vectors and returns a number, changing sign under the exchange of any two arguments. A $0$-form is a function. A $1$-form is a covector field; the coordinate differentials $\dd{\xi^\alpha}$ form a basis, dual to the coordinate vector fields $\partial_\alpha$ in the sense $\dd{\xi^\alpha}(\partial_\beta) = \delta^\alpha_\beta$, so a general one-form is $\lambda = \lambda_\alpha\, \dd{\xi^\alpha}$. Higher forms are built with the \emph{wedge product}, the antisymmetrized tensor product; for one-forms,
\begin{flalign}
    (\lambda \wedge \mu)(X, Y) = \lambda(X)\mu(Y) - \lambda(Y)\mu(X),
\end{flalign}
so $\lambda\wedge\mu = -\mu\wedge\lambda$ and in particular $\dd{\xi^\alpha}\wedge\dd{\xi^\alpha} = 0$ (no sum). A general $k$-form is a combination $\omega = \frac{1}{k!}\,\omega_{\alpha_1\cdots\alpha_k}\, \dd{\xi^{\alpha_1}}\wedge\cdots\wedge\dd{\xi^{\alpha_k}}$ with totally antisymmetric components. On a $d$-dimensional manifold, $k$-forms with $k > d$ vanish identically (antisymmetry in more slots than there are dimensions), and the top forms, $k = d$, are one-dimensional at each point: they are the objects that can be integrated over $M$, i.e. volume elements. That's the ultimate reason forms are the right language for us: Liouville's theorem is a statement about a volume, and the volume in question will be built by wedging the symplectic form with itself.

Why antisymmetry, though? Because $k$-forms are the machines that measure \emph{oriented $k$-dimensional areas} spanned by their arguments, and oriented areas flip sign when two spanning vectors are exchanged. A $2$-form assigns to a pair of vectors the (weighted, oriented) area of the parallelogram they span. Keep that picture; the symplectic form is exactly such an area-measurer, and ``symplectic geometry is the geometry of area'' is a slogan we'll be able to cash out precisely in \S\ref{sec:symptrans}.

\subsection{The exterior derivative}

The \emph{exterior derivative} $\dd{}$ takes $k$-forms to $(k+1)$-forms. On functions it's the differential, $\dd{f} = \partial_\alpha f\, \dd{\xi^\alpha}$. On a general form written in coordinates it differentiates the components and wedges on the new differential:
\begin{flalign}
    \dd{\br{\tfrac{1}{k!}\,\omega_{\alpha_1\cdots\alpha_k}\, \dd{\xi^{\alpha_1}}\wedge\cdots\wedge\dd{\xi^{\alpha_k}}}}
    = \tfrac{1}{k!}\,\partial_\beta\, \omega_{\alpha_1\cdots\alpha_k}\; \dd{\xi^{\beta}}\wedge \dd{\xi^{\alpha_1}}\wedge\cdots\wedge\dd{\xi^{\alpha_k}}. \label{extd}
\end{flalign}
The single most important property:
\begin{empheq}[box=\mathbox]{align}
    \dd{}^2 = 0, \label{dd0}
\end{empheq}
for the reason that $\dd{}^2$ produces $\partial_\beta\partial_\gamma\,\omega_{\cdots}\, \dd{\xi^\gamma}\wedge\dd{\xi^\beta}\wedge\cdots$, a symmetric object (second partials) contracted with an antisymmetric one ($\dd{\xi^\gamma}\wedge\dd{\xi^\beta}$), which vanishes. A form $\omega$ with $\dd{\omega} = 0$ is called \emph{closed}; a form of the shape $\omega = \dd{\theta}$ is called \emph{exact}. By \eqref{dd0} every exact form is closed. The converse is the Poincar\'e lemma: \emph{locally} (on any contractible patch) every closed form is exact. We've already met this dichotomy in disguise: condition \eqref{integrability} says exactly that the one-form $\lambda = \lambda_\alpha \dd{\xi^\alpha}$ is closed, since
\begin{flalign}
    \dd{\lambda} = \partial_\beta \lambda_\alpha\, \dd{\xi^\beta}\wedge\dd{\xi^\alpha} = \tfrac{1}{2}\br{\partial_\beta\lambda_\alpha - \partial_\alpha\lambda_\beta}\, \dd{\xi^\beta}\wedge\dd{\xi^\alpha},
\end{flalign}
and being a gradient, $\lambda = \dd{G}$, says it's exact. The discussion of locally versus globally Hamiltonian fields in \S\ref{sec:poisson} was secretly the closed-vs-exact distinction, and we'll say it in those words from now on.

\subsection{Interior product, pullback, Lie derivative}

Three more operations and the kit is complete.

The \emph{interior product} (or contraction) $i_X$ with a vector field $X$ takes $k$-forms to $(k-1)$-forms by feeding $X$ into the first slot:
\begin{flalign}
    (i_X\,\omega)(Y_1, \dots, Y_{k-1}) = \omega(X, Y_1, \dots, Y_{k-1}).
\end{flalign}
On functions $i_X f \equiv 0$; on one-forms $i_X\lambda = \lambda(X) = \lambda_\alpha X^\alpha$, a function. It's an antiderivation: $i_X(\lambda\wedge\mu) = (i_X\lambda)\wedge\mu + (-1)^{\deg\lambda}\,\lambda\wedge(i_X\mu)$, and $i_X i_X = 0$ (antisymmetry of the form's slots).

The \emph{pullback}: any smooth map $\varphi : M \to N$ drags forms on $N$ back to forms on $M$ by pre-composing with the derivative of $\varphi$,
\begin{flalign}
    (\varphi^*\omega)\big|_x\,(X_1, \dots, X_k) = \omega\big|_{\varphi(x)}\br{\varphi_* X_1, \dots, \varphi_* X_k},
\end{flalign}
where $\varphi_*$ is the Jacobian map on tangent vectors. In coordinates, pullback is ``substitute and use the chain rule'', which is why it's the natural formalization of ``how does this object look after a change of variables/after being carried along a flow''. Pullback commutes with everything in sight: $\varphi^*(\lambda\wedge\mu) = \varphi^*\lambda \wedge \varphi^*\mu$ and, crucially, $\varphi^*\,\dd{\omega} = \dd{(\varphi^*\omega)}$.

The \emph{Lie derivative} $\Lie_X$ measures the rate of change of a geometric object as it's dragged along the flow of $X$ (we'll construct flows carefully in \S\ref{sec:flows}; for now, $\varphi_s$ is the map ``follow the trajectories of $X$ for time $s$''):
\begin{flalign}
    \Lie_X\, \omega \equiv \dv{s}\bigg|_{s=0} \varphi_s^*\,\omega. \label{lieder}
\end{flalign}
The statement ``$\omega$ is invariant under the flow of $X$'' is exactly $\Lie_X\omega = 0$. Computing \eqref{lieder} from the definition is painful; mercifully there is \emph{Cartan's magic formula},
\begin{empheq}[box=\mathbox]{align}
    \Lie_X = i_X \circ \dd{} + \dd{} \circ i_X, \label{cartan}
\end{empheq}
valid on forms of any degree. (Sanity check on functions: $\Lie_X f = i_X\,\dd{f} + 0 = X^\alpha\partial_\alpha f$, the directional derivative, as it must be.) We won't prove \eqref{cartan} in general (induction on form degree, using that both sides are derivations agreeing on functions and differentials), but we will use it constantly, because it converts flow-invariance questions into pure algebra with $\dd{}$ and $i_X$. Watch how much work it does in \S\ref{sec:symptrans}.

Finally, for the record and for orientation: forms are the natural integrands ($k$-forms integrate over $k$-dimensional oriented surfaces), and the exterior derivative is dual to the boundary operator via \emph{Stokes' theorem}, $\int_S \dd{\omega} = \int_{\partial S}\omega$, which contains the divergence and curl theorems of vector calculus as special cases. We'll need it exactly once (for integral invariants, \S\ref{sec:symptrans}), and we'll cite it rather than prove it.

\section{The Symplectic Form}\label{sec:symplectic}

\subsection{From the symplectic matrix to the symplectic form}

Return to the matrix $\tilde{\Omega}^{\alpha\beta}$ of \eqref{hameqs-xi} and its inverse $\tilde{\Omega}_{\alpha\beta}$ (numerically $\tilde\Omega^{-1} = -\tilde\Omega = \tilde\Omega^{\mathrm{T}}$ in canonical coordinates, by \eqref{omega-props}). How do these components behave under a change of phase-space coordinates $\xi \to \tilde{\xi}(\xi)$? The bracket \eqref{PB} is built from two gradients, each transforming with an inverse Jacobian, so demanding that $\cbr{f,g}$ be a coordinate-independent object forces the upper-index components to transform with two Jacobians,
\begin{flalign}
    \tilde{\Omega}^{\alpha\beta} \longrightarrow \pdv{\tilde{\xi}^\alpha}{\xi^\mu}\, \pdv{\tilde{\xi}^\beta}{\xi^\nu}\, \tilde{\Omega}^{\mu\nu},
\end{flalign}
i.e. $\tilde\Omega^{\alpha\beta}$ are the components of an antisymmetric contravariant $2$-tensor (a \emph{bivector}), and its inverse $\tilde{\Omega}_{\alpha\beta}$ transforms with two inverse Jacobians: the components of an antisymmetric covariant $2$-tensor, which is to say, \emph{a $2$-form}. That $2$-form is the fundamental object of the subject. In canonical coordinates,
\begin{empheq}[box=\mathbox]{align}
    \Omega \equiv \frac{1}{2}\,\tilde{\Omega}_{\alpha\beta}\;\dd{\xi^\alpha}\wedge\dd{\xi^\beta} = \dd{p_a}\wedge\dd{q^a}, \label{sympform}
\end{empheq}
the \emph{symplectic form} on phase space. (The evaluation: with the ordering \eqref{xidef} and $\tilde\Omega$ as in \eqref{hameqs-xi}, the inverse matrix has $\tilde{\Omega}_{a,\,n+b} = -\delta_{ab} = -\tilde{\Omega}_{n+b,\,a}$, and the half-sum over both index orders assembles into $\sum_a \dd{p_a}\wedge\dd{q^a}$; the sign conventions are chosen so that \eqref{sympform} together with \eqref{PB} reproduces $\cbr{q^a, p_b} = +\delta^a_b$. Some texts write $\Omega = \dd{q^a}\wedge\dd{p_a}$, the opposite sign; this is the same convention fork flagged in the box in \S\ref{sec:poisson}, wearing different clothes.)

The symplectic form has two defining properties, both inherited from the matrix:
\begin{enumerate}[label=(\roman*)]
    \item \textbf{Closedness:} $\dd{\Omega} = 0$. In canonical coordinates the components are constant, so this is immediate from \eqref{extd}; but the statement is coordinate-independent and, as we'll see in \S\ref{sec:sympman}, it's the property that carries the entire dynamical content (it will be equivalent to the Jacobi identity).
    \item \textbf{Nondegeneracy:} the map $X \mapsto i_X\Omega$ from vector fields to one-forms is invertible; equivalently, $\Omega(X, Y) = 0$ for all $Y$ forces $X = 0$; equivalently, $\det \tilde{\Omega}_{\alpha\beta} \neq 0$. In canonical coordinates $\det\tilde{\Omega} = 1$.
\end{enumerate}

Nondegeneracy deserves a paragraph, because it's the exact structural analogue of something familiar. A Riemannian metric $g_{ab}$ is a \emph{symmetric} nondegenerate $2$-tensor, and its nondegeneracy gives the musical isomorphism between vectors and covectors: $X^a \mapsto g_{ab}X^b$, ``lowering the index'', with inverse $g^{ab}$. The symplectic form is the \emph{antisymmetric} counterpart: nondegeneracy gives an isomorphism
\begin{flalign}
    X \longmapsto i_X \Omega, \qquad \text{in components } X^\alpha \mapsto \tilde{\Omega}_{\beta\alpha} X^\alpha,
\end{flalign}
between vector fields and one-forms, with inverse supplied by $\tilde{\Omega}^{\alpha\beta}$. Metric geometry converts vectors to covectors symmetrically; symplectic geometry converts them antisymmetrically. Practically, this isomorphism is the machine that will let us convert the one-form $\dd{H}$ (which every function has) into a vector field (which a dynamics needs). That conversion \emph{is} Hamilton's equations, and it's one section away.

As a warm-up with the machinery, contract $\Omega$ with a coordinate vector field, say $\partial/\partial q^b$:
\begin{flalign}
    i_{\partial/\partial q^b}\,\Omega = i_{\partial/\partial q^b}\br{\dd{p_a}\wedge\dd{q^a}}
    = \underbrace{\dd{p_a}\br{\tfrac{\partial}{\partial q^b}}}_{=\,0}\dd{q^a} - \dd{p_a}\, \underbrace{\dd{q^a}\br{\tfrac{\partial}{\partial q^b}}}_{=\,\delta^a_b} = -\dd{p_b},
\end{flalign}
and symmetrically $i_{\partial/\partial p_b}\Omega = \dd{q^b}$. Moving in a coordinate direction on phase space corresponds, under the symplectic isomorphism, to the differential of the \emph{conjugate} coordinate (with a sign). The pairing of $q$'s with $p$'s is built into $\Omega$'s bones.

\subsection{The tautological one-form: why $T^*\mathcal{Q}$ is symplectic for free}

In canonical coordinates, \eqref{sympform} visibly says $\Omega = \dd{\theta}$ for the one-form
\begin{flalign}
    \theta = p_a\, \dd{q^a}, \label{taut}
\end{flalign}
(so $\Omega$ is not merely closed but exact on a cotangent bundle). The form \eqref{taut} looks coordinate-bound, but it is secretly intrinsic, and Arnold's construction of it is one of the slickest small arguments in the subject.

\textbf{Theorem.} The cotangent bundle $\Gamma = T^*\mathcal{Q}$ carries a canonical one-form $\theta$, defined without any choice of coordinates, whose coordinate expression in any chart of the form $(q^a, p_a)$ is \eqref{taut}.

\emph{Proof (construction).} A point of $\Gamma$ is a pair $x = (q, p)$: a configuration $q\in\mathcal{Q}$ and a covector $p \in T^*_q\mathcal{Q}$. There is a natural projection $\pi : \Gamma \to \mathcal{Q}$, $(q,p)\mapsto q$, which simply forgets the momentum, and its derivative $\pi_* : T_x\Gamma \to T_q\mathcal{Q}$ pushes tangent vectors of phase space down to tangent vectors of configuration space. Now take any tangent vector $\zeta \in T_x\Gamma$ and define
\begin{empheq}[box=\mathbox]{align}
    \theta\big|_x(\zeta) \equiv p\br{\pi_*\, \zeta}. \label{taut-intrinsic}
\end{empheq}
Read it out loud: to evaluate $\theta$ on a phase-space direction $\zeta$ at the point $(q,p)$, project $\zeta$ down to a configuration-space direction and feed it to \emph{the momentum covector that is the point itself}. The point of phase space is a covector; the tautological form just \emph{uses} it. No structure beyond the bundle itself is involved, hence ``tautological'' (also ``canonical one-form'' or ``Liouville form'' in the literature).

In coordinates: $\zeta = \dot{q}^a\,\partial_{q^a} + \dot{p}_a\,\partial_{p_a}$ projects to $\pi_*\zeta = \dot{q}^a\,\partial_{q^a}$, and $p(\pi_*\zeta) = p_a \dot{q}^a = (p_a\,\dd{q^a})(\zeta)$. So $\theta = p_a\,\dd{q^a}$ in \emph{every} chart induced from configuration-space coordinates, which is the claim. $\blacksquare$

Consequently
\begin{empheq}[box=\mathbox]{align}
    \Omega = \dd{\theta} = \dd{p_a}\wedge\dd{q^a}
\end{empheq}
is likewise canonical: \emph{every} cotangent bundle is a symplectic manifold, with no choices made. This is the precise sense in which phase space has more structure than velocity space: $T^*\mathcal{Q}$ comes symplectic from the factory, while $T\mathcal{Q}$ acquires such structure only after a Lagrangian is chosen to identify it with $T^*\mathcal{Q}$ (via \eqref{canmom}). And notice which physical object $\theta$ is: $p_a\,\dd{q^a}$ is the integrand of the abbreviated action $\int p\,\dd{q}$, the quantity whose stationarity gives Maupertuis' principle and whose quantization gives Bohr--Sommerfeld. The tautological form is not a formal device; it's the action integrand, promoted to geometry.


\section{Hamilton's Equations Without Coordinates}\label{sec:coordfree}

We now have all the pieces to state the dynamics geometrically. A Hamiltonian system is a triple: a phase space $\Gamma$, a symplectic form $\Omega$ on it, and a function $H : \Gamma \to \mathbb{R}$. The dynamics is the flow of the vector field $X_H$ determined by
\begin{empheq}[box=\mathbox]{align}
    i_{X_H}\,\Omega = -\dd{H}. \label{coordfree}
\end{empheq}
That single equation is Hamilton's equations. Unpack it in canonical coordinates to check. Write $X_H = A^a\,\partial_{q^a} + B_a\,\partial_{p_a}$ for unknown components. Then, using the warm-up contractions of \S\ref{sec:symplectic} and linearity,
\begin{flalign}
    i_{X_H}\Omega = A^a\br{i_{\partial_{q^a}}\Omega} + B_a\br{i_{\partial_{p_a}}\Omega} = -A^a\, \dd{p_a} + B_a\, \dd{q^a},
\end{flalign}
while
\begin{flalign}
    -\dd{H} = -\pdv{H}{q^a}\,\dd{q^a} - \pdv{H}{p_a}\,\dd{p_a}.
\end{flalign}
Matching coefficients of the independent basis forms $\dd{q^a}$ and $\dd{p_a}$:
\begin{flalign}
    A^a = \pdv{H}{p_a}, \qquad B_a = -\pdv{H}{q^a},
\end{flalign}
so the integral curves of $X_H$ (curves whose velocity is $X_H$, i.e. $\dot{q}^a = A^a$, $\dot{p}_a = B_a$) satisfy precisely \eqref{hameqs}. Note the role of each hypothesis: \emph{nondegeneracy} of $\Omega$ is what guarantees that \eqref{coordfree} has exactly one solution $X_H$ for a given $H$, i.e. that a Hamiltonian determines a dynamics; \emph{closedness} hasn't been used yet, and its moment comes in \S\ref{sec:symptrans}, where it will be exactly the statement that the resulting flow preserves $\Omega$.

(On the sign in \eqref{coordfree}: with our $\Omega = \dd{p_a}\wedge\dd{q^a}$, the minus sign is what lands on \eqref{hameqs} with the physicists' signs. Texts that take $\Omega = \dd{q^a}\wedge\dd{p_a}$ write $i_{X_H}\Omega = +\dd{H}$; the two sign flips cancel and the physics agrees. Same fork, third costume.)

Energy conservation, third derivation, now purely algebraic: the rate of change of $H$ along its own flow is
\begin{flalign}
    \Lie_{X_H} H = \dd{H}(X_H) = -\br{i_{X_H}\Omega}(X_H) = -\Omega(X_H, X_H) = 0, \label{econs-geo}
\end{flalign}
where the last step is nothing but the antisymmetry of $\Omega$: any $2$-form vanishes when fed the same vector twice. Compare the coordinate computation \eqref{Hdot}: what looked there like a fortunate cancellation of two sums is here the single word ``antisymmetry''. That's the geometry paying rent. And notice what the argument \emph{didn't} use: nothing about $H$, nothing about coordinates, not even closedness of $\Omega$. Conservation of the generator is a structural fact about antisymmetric dynamics, full stop. (We'll upgrade this observation into a two-line Noether-type theorem in \S\ref{sec:hamvf}.)

The picture to keep: $\dd{H}$ is a covector field, pointing ``uphill'' across the level sets of $H$. The symplectic isomorphism converts it into the vector field $X_H$, and by \eqref{econs-geo} that vector field is everywhere \emph{tangent} to the level sets. The conversion rotates uphill into along-the-contour. This is the coordinate-free version of the ``sideways gradient flow'' picture of \S\ref{sec:phasespace}, and it's exact, not a cartoon.

\section{Flows}\label{sec:flows}

\subsection{From vector fields to flows}

A vector field $X$ on a manifold assigns a velocity to every point; solving the ODE $\dot{\xi} = X(\xi)$ from every initial condition simultaneously produces the \emph{flow} of $X$: the family of maps
\begin{flalign}
    \varphi^X_s : \Gamma \to \Gamma, \qquad \varphi^X_s(\xi_0) = \text{(solution at parameter $s$ starting from $\xi_0$)},
\end{flalign}
which slides every point of the manifold along its trajectory for parameter-time $s$. Uniqueness of ODE solutions gives the defining algebraic property, the \emph{group law}
\begin{flalign}
    \varphi^X_s \circ \varphi^X_u = \varphi^X_{s+u}, \qquad \varphi^X_0 = \mathrm{id}, \qquad \br{\varphi^X_s}^{-1} = \varphi^X_{-s}: \label{grouplaw}
\end{flalign}
flowing for time $u$ and then for time $s$ is flowing for time $s + u$. A flow is a one-parameter group of transformations of the manifold, and the vector field is its \emph{generator}: the derivative of the family at $s = 0$. (Strictly, completeness of the flow for all $s$ needs mild hypotheses; for the compact level sets typical of confining Hamiltonians it's automatic, and we won't fuss.)

How does an observable $f$ change under the flow? Define $f_s \equiv f \circ \varphi^X_s$ (the pullback $\varphi_s^{X\,*}f$). Then
\begin{flalign}
    \dv{s} f_s = \Lie_X f_s = X^\alpha\, \partial_\alpha f_s,
\end{flalign}
a linear ODE on the space of observables, solved formally by exponentiation:
\begin{flalign}
    f_s = e^{s \Lie_X} f, \qquad \text{i.e.} \qquad \varphi_s^{X\,*} = e^{s\Lie_X} \label{expflow}
\end{flalign}
(convergence for analytic data; formally always, which is the register in which we'll use it). If the generator itself carries explicit parameter dependence, the correct statement acquires the by-now-familiar ordering caveats, and one writes $\varphi^{X\,*}_s = e^{s(\Lie_X + \partial_s)}$ in the sense of the parameter-ordered exponential; we'll only need the autonomous case, but the pattern, evolution $=$ exponentiated generator, with ordering subtleties when the generator moves, is one we've seen before in these notes' companion volume, wearing the costume $T(t;t_0) = e^{-iH(t-t_0)/\hbar}$ and its time-ordered generalization. It is the same mathematics.

The simplest example, worth doing in full because of who the generator turns out to be: on $\Gamma = \mathbb{R}^2$ with coordinates $(q, p)$, consider $X = \partial_q$ (unit velocity in $q$, none in $p$). The ODE is $\dot{q} = 1$, $\dot{p} = 0$, so $\varphi^X_s(q, p) = (q + s, p)$: the flow is rigid translation in position. And by Taylor's theorem, $f(q+s, p) = e^{s\,\partial_q}f(q,p)$, which is \eqref{expflow} in its most familiar clothing. Now ask: is this flow Hamiltonian, and if so, what generates it? By \eqref{coordfree} we need $i_{\partial_q}\Omega = -\dd{G}$, and we computed $i_{\partial_q}\Omega = -\dd{p}$ in \S\ref{sec:symplectic}. So $G = p$:
\begin{empheq}[box=\mathbox]{align}
    \text{\emph{momentum generates translations},} \qquad \varphi_s(q,p) = (q+s,\,p), \quad G = p.
\end{empheq}
Likewise $i_{\partial_p}\Omega = \dd{q} = -\dd{(-q)}$, so $-q$ generates translations in momentum; and the reader should check that the Hamiltonian of the free particle, $H = p^2/2m$, generates $\varphi_t(q,p) = (q + pt/m,\, p)$, uniform straight-line motion, as it had better. The slogan ``momentum generates translations'', usually first met as a quantum fact about $e^{-ia\hat{p}/\hbar}$, is a classical fact about the symplectic form, and the quantum version is its faithful image.

\section{Hamiltonian Vector Fields and the Algebra of Observables}\label{sec:hamvf}

\subsection{Every observable generates a flow}

Nothing in the construction \eqref{coordfree} was special to ``the'' Hamiltonian: any observable $f$ defines its \emph{Hamiltonian vector field}
\begin{empheq}[box=\mathbox]{align}
    i_{X_f}\,\Omega = -\dd{f}, \qquad X_f = \tilde{\Omega}^{\alpha\beta}\,\partial_\beta f\; \partial_\alpha \;=\; \cbr{\,\cdot\,, f}\text{ as an operator}, \label{hamvf}
\end{empheq}
where the middle expression solves the defining equation in canonical coordinates (invert the isomorphism with $\tilde{\Omega}^{\alpha\beta}$) and the last equality is checked by acting on an arbitrary $g$: $X_f\, g = \tilde{\Omega}^{\alpha\beta}\partial_\beta f\, \partial_\alpha g = \cbr{g, f}$. So the derivation $\cbr{\,\cdot\,, f}$ promised in \S\ref{sec:poisson} is a bona fide vector field, and every observable is simultaneously a quantity (a number on each state) and a generator (of a flow $\varphi^f_s \equiv \varphi^{X_f}_s$ on phase space). This dual citizenship of observables, number-valued and transformation-generating at once, is one of the deepest features Hamiltonian mechanics bequeaths to quantum mechanics, where every Hermitian operator is likewise both a measurable and the generator of a unitary flow.

The bracket now acquires its second, flow-theoretic characterization:
\begin{flalign}
    \cbr{g, f} = X_f\, g = \dv{s}\bigg|_{s=0}\; g \circ \varphi^{f}_s\,, \label{PB-flow}
\end{flalign}
\emph{the Poisson bracket of $g$ with $f$ is the rate of change of $g$ along the flow generated by $f$}. (This is the definition Arnold takes as primary, with his sign convention; see the box in \S\ref{sec:poisson}.) Evolution \eqref{PB-evolution} is the special case $f = H$: dynamics is just the flow generated by the energy, on the same footing as translations are the flow generated by momentum. Time evolution loses its privileged status and becomes one flow among many; that demotion is one of the most useful conceptual moves in the subject.

Which flows arise this way? Exactly the question settled in \S\ref{sec:poisson}, which we can now phrase in form language and answer in one line. Given a vector field $X$, it is of the form $X_f$ iff the one-form $-i_X\Omega$ is exact ($= \dd{f}$); it is \emph{locally} of that form iff $-i_X\Omega$ is closed,
\begin{flalign}
    \dd{\br{i_X\,\Omega}} = 0 \quad \Longleftrightarrow \quad X \text{ locally Hamiltonian}, \label{locham}
\end{flalign}
the Poincar\'e lemma bridging the two as always; and the coordinate condition \eqref{integrability} was nothing but \eqref{locham} in components. The obstruction to ``locally'' upgrading to ``globally'' is the first de Rham cohomology of phase space, as promised\footnote{For the cylinder $\Gamma = S^1\times\mathbb{R}$ of the pendulum, the closed-but-not-exact one-form $\dd{\theta}$ (the angle differential, closed since locally $\theta$ is a function, non-exact since $\theta$ isn't one globally) corresponds to the locally Hamiltonian vector field $-\partial_p$: uniform drift in angular momentum, generated locally by the angle, which fails to exist as a single-valued observable. Topology of phase space $=$ existence of would-be generators. It's a real effect, not a pathology.}.

\subsection{The Lie algebra homomorphism}

We have two bracket-like operations in play: the Poisson bracket of observables, and the commutator of vector fields, $[X, Y] = XY - YX$ (which is again a vector field, by the same cancellation of second derivatives used in the Jacobi proof of \S\ref{sec:poisson}). The central structural theorem of the section is that the map $f \mapsto X_f$ intertwines them.

\textbf{Theorem.} For any observables $f, g$:
\begin{empheq}[box=\mathbox]{align}
    \sbr{X_f,\, X_g} = X_{\cbr{g,\,f}}. \label{liehom}
\end{empheq}

\emph{Proof.} Act with both sides on an arbitrary test observable $h$ and use \eqref{hamvf} three times:
\begin{flalign}
    \sbr{X_f, X_g}\, h &= X_f\br{X_g h} - X_g\br{X_f h} = X_f \cbr{h, g} - X_g\cbr{h, f} \notag \\
    &= \cbr{\cbr{h,g}, f} - \cbr{\cbr{h,f}, g}.
\end{flalign}
Now use the Jacobi identity, arranged as $\{\{h,g\},f\} + \{\{g,f\},h\} + \{\{f,h\},g\} = 0$, to rearrange the difference:
\begin{flalign}
    \cbr{\cbr{h,g},f} - \cbr{\cbr{h,f},g} &= \cbr{\cbr{h,g},f} + \cbr{\cbr{f,h},g} \notag \\
    &= -\cbr{\cbr{g,f}, h} = \cbr{h, \cbr{g,f}} = X_{\cbr{g,f}}\, h,
\end{flalign}
and since $h$ was arbitrary, \eqref{liehom} follows. $\blacksquare$

(Mind the order on the right of \eqref{liehom}: with our sign conventions it's $\cbr{g,f}$, not $\cbr{f,g}$. Chasing that sign through is a rite of passage; in Arnold's conventions the analogous statement carries the opposite arrangement, and both are correct in their own books. The box in \S\ref{sec:poisson} was not kidding.)

The content of \eqref{liehom}: the observables form a Lie algebra under $\cbr{\cdot,\cdot}$ (antisymmetry, bilinearity, Jacobi: \S\ref{sec:poisson}), the vector fields form a Lie algebra under the commutator, and $f \mapsto X_{f}$ is (up to the sign bookkeeping) a \emph{homomorphism of Lie algebras} between them. Its kernel is the set of observables with $X_f = 0$, i.e. $\dd{f} = 0$ by nondegeneracy: the locally constant functions. So observables that differ by a constant generate identical flows, and no further ambiguity exists; the algebra of observables is a one-dimensional central extension of the algebra of (globally) Hamiltonian vector fields. That sentence, suitably quantized, is why the quantum phase $e^{i\alpha}$ is invisible: states related by constants-in-the-generator differ by nothing observable. The classical seed of the ray structure of quantum mechanics is already here.

\subsection{Symmetries and conservation: a two-line Noether theorem}

The flow characterization \eqref{PB-flow} makes the symmetry--conservation link almost embarrassingly short. Say an observable $g$ is a \emph{symmetry generator} of the dynamics if $H$ is invariant under the flow of $g$, i.e. $\Lie_{X_g}H = \cbr{H, g} = 0$. Say $g$ is \emph{conserved} if it's constant along the dynamics, i.e. $\cbr{g, H} = 0$. But by antisymmetry these are \emph{the same equation}:
\begin{empheq}[box=\mathbox]{align}
    \cbr{H, g} = 0 \quad&\Longleftrightarrow\quad \cbr{g, H} = 0 : \label{noether} \\
    \text{$g$ generates a symmetry of $H$} \quad&\iff\quad \text{$g$ is conserved.} \notag
\end{empheq}
Momentum is conserved when $H$ is translation-invariant, and $H$ is translation-invariant when momentum is conserved, and these are one fact, not two facts joined by a theorem. Energy conservation itself is the diagonal case $g = H$. (The Lagrangian Noether theorem, with its variational hypotheses and boundary terms, is the same statement met under worse lighting; on phase space the symmetry--conservation duality is just the antisymmetry of $\Omega$.)


\section{Symplectic Manifolds}\label{sec:sympman}

\subsection{The abstraction}

Everything from \S\ref{sec:symplectic} onward used exactly two properties of $\Omega$, closedness and nondegeneracy, and nothing else about the cotangent-bundle origin of phase space. So abstract:

\begin{empheq}[box=\mathbox]{align*}
    &\text{A \emph{symplectic manifold} is a pair } (\mathcal{M}, \Omega): \\
    &\text{a smooth manifold equipped with a closed, nondegenerate } 2\text{-form.}
\end{empheq}

Any such pair supports the entire apparatus we've built: Hamiltonian vector fields via \eqref{hamvf}, Poisson brackets via $\cbr{f,g} = \Omega(X_g, X_f)$ (check in coordinates that this reproduces \eqref{PB}), the evolution law, the Lie algebra homomorphism, and, as we're about to show, canonical transformations and Liouville's theorem. Mechanics runs wherever $(\mathcal{M},\Omega)$ exists.

The examples to keep in mind:
\begin{enumerate}[label=(\roman*)]
    \item $\mathbb{R}^{2n}$ with $\Omega = \dd{p_a}\wedge\dd{q^a}$: the local model, and by Darboux's theorem below, in a precise sense the \emph{only} local model.
    \item Any cotangent bundle $T^*\mathcal{Q}$ with $\Omega = \dd{\theta}$: \S\ref{sec:symplectic}. Here $\Omega$ is exact, not merely closed.
    \item The sphere $S^2$ with $\Omega = $ its area form (in spherical coordinates, $\sin\vartheta\,\dd{\vartheta}\wedge\dd{\varphi}$). Nondegenerate because the area form of any oriented surface is (a $2$-form on a $2$-manifold is nondegenerate iff it's nowhere zero); closed because \emph{every} top-degree form is closed (there are no $3$-forms on a $2$-manifold to be equal to). This is the phase space of a classical spin: the observables $(S_x, S_y, S_z)$, the Cartesian embedding functions scaled by the spin magnitude, have Poisson brackets $\cbr{S_x, S_y} = S_z$ and cyclic, and Hamiltonians built from them generate rigid rotations of the sphere. Note $S^2$ is \emph{not} a cotangent bundle of anything: it's compact, and cotangent bundles never are\footnote{Quick reason: on $T^*\mathcal{Q}$ the symplectic form is exact, $\Omega = \dd{\theta}$; on a compact $\mathcal{M}$ without boundary, exactness would give $\int_{\mathcal{M}}\Omega^n = \int_{\mathcal{M}}\dd{(\theta\wedge\Omega^{n-1})} = 0$ by Stokes, contradicting the fact that $\Omega^n$ is a volume form with nonzero total volume. So compact symplectic manifolds have closed-but-not-exact $\Omega$, and genuinely enlarge the theory beyond cotangent bundles.}. Spin systems are the standard reminder that ``phase space $= T^*(\text{configuration space})$'' is a common case, not a law.
\end{enumerate}

\subsection{Symplectic manifolds are even-dimensional}

Not every manifold can carry a symplectic form. The first obstruction is linear-algebraic:

\textbf{Proposition.} A symplectic manifold has even dimension.

\emph{Proof.} At any point, $\Omega$ has an antisymmetric component matrix $\Omega_{\alpha\beta}$ with $\det\Omega_{\alpha\beta}\neq 0$ (nondegeneracy). Antisymmetry says $\Omega^{\mathrm{T}} = -\Omega$ as matrices, so
\begin{flalign}
    \det\Omega = \det\Omega^{\mathrm{T}} = \det(-\Omega) = (-1)^{d}\det\Omega,
\end{flalign}
with $d = \dim\mathcal{M}$. If $d$ is odd this forces $\det\Omega = -\det\Omega = 0$, contradicting nondegeneracy. $\blacksquare$

(Odd-dimensional spaces aren't exiled from mechanics; they host the \emph{contact} geometry of constant-energy surfaces and time-extended phase space, a sibling theory we won't develop.)

\subsection{Darboux's theorem: no local invariants}

Here is the theorem that fixes the character of the whole subject.

\textbf{Theorem (Darboux).} Let $(\mathcal{M}, \Omega)$ be a symplectic manifold of dimension $2n$. Around every point there exist local coordinates $(q^1,\dots,q^n,p_1,\dots,p_n)$ in which
\begin{flalign}
    \Omega = \dd{p_a}\wedge\dd{q^a} \qquad \text{with \emph{constant} coefficients, exactly.} \label{darboux}
\end{flalign}

Pause on what's being claimed, because it's stronger than it may look. For a Riemannian metric, the analogous statement is \emph{false}: one can arrange $g_{ab} = \delta_{ab}$ \emph{at a point} (normal coordinates), but not on a neighbourhood; the obstruction is the curvature tensor, a genuine local invariant that distinguishes a sphere from a plane at arbitrarily small scales. Darboux says symplectic geometry has \emph{no curvature, no local invariants at all}: every symplectic manifold of a given dimension is locally identical to every other, and locally identical to the standard $\mathbb{R}^{2n}$. All the information in a symplectic manifold is global (topology, total volumes, and subtler invariants of that kind). This is why the mechanics of \S\S\ref{sec:hamilton}--\ref{sec:poisson}, developed in coordinates on $\mathbb{R}^{2n}$, was secretly fully general: Darboux coordinates make any symplectic manifold look like that, one chart at a time. Canonical coordinates aren't a convenience; they're a theorem.

\emph{Proof sketch (Arnold's induction).} The strategy is induction on $n$, constructing the coordinate pair $(q^1, p_1)$ by hand and splitting it off. Take any function $p_1$ with $\dd{p_1}\neq 0$ near the point (a coordinate function will do). Its Hamiltonian field $X_{p_1}$ is nonvanishing there, so we may choose a function $q^1$ with $X_{p_1}q^1 = 1$, i.e. $\cbr{q^1, p_1} = 1$ (solve the ODE ``$q^1$ increases at unit rate along the flow of $p_1$'', taking $q^1 = 0$ on a hypersurface transverse to the flow; existence and smooth dependence come from the standard straightening theorem for nonvanishing vector fields). The two flows $\varphi^{q^1}_s$ and $\varphi^{p_1}_u$ commute, because by \eqref{liehom} $\sbr{X_{q^1}, X_{p_1}} = X_{\cbr{p_1, q^1}} = X_{-1} = 0$, constants generating nothing. Using the two commuting flows as two of the coordinate directions, one shows the remaining directions, the joint level set of $(q^1, p_1)$ swept out suitably, form a $(2n-2)$-dimensional manifold on which $\Omega$ restricts to a closed nondegenerate form; apply the inductive hypothesis there, and verify that the assembled coordinates give exactly \eqref{darboux} (the cross terms vanish precisely because of the commutation and the normalizations $\cbr{q^1,p_1}=1$). The base case $n=0$ is empty. $\blacksquare$

We've been honest that this is a sketch; the full bookkeeping (that the restricted form is symplectic, that no cross terms survive) is a page of careful checking in Arnold \S43. The two ideas worth retaining are (i) coordinates are \emph{built from dynamics}: the canonical pair $(q^1, p_1)$ is manufactured out of a flow and its clock, and (ii) the Jacobi identity, via \eqref{liehom}, is what makes the construction close, which is the promised sense in which $\dd{\Omega} = 0$ (equivalent to Jacobi, as one can check by expressing the Jacobiator of \eqref{PB} in terms of $\dd{\Omega}$) carries the dynamical content of the subject.

\section{Symplectic (Canonical) Transformations}\label{sec:symptrans}

\subsection{Definition and the classical propositions}

A \emph{symplectic transformation} (the geometer's name) or \emph{canonical transformation} (the physicist's) is a smooth invertible map $\varphi : \mathcal{M} \to \mathcal{M}$ preserving the symplectic form under pullback:
\begin{empheq}[box=\mathbox]{align}
    \varphi^*\,\Omega = \Omega. \label{symptrans-def}
\end{empheq}
In coordinates, writing the Jacobian $\Lambda^\mu{}_\alpha = \pdv*{\tilde{\xi}^\mu}{\xi^\alpha}$, condition \eqref{symptrans-def} reads $\Lambda^{\mathrm{T}}\tilde{\Omega}^{-1}\Lambda = \tilde{\Omega}^{-1}$, or equivalently for the bracket matrix,
\begin{flalign}
    \Lambda\,\tilde{\Omega}\,\Lambda^{\mathrm{T}} = \tilde{\Omega}, \label{symp-matrix}
\end{flalign}
which for \emph{linear} maps defines the symplectic group $Sp(2n,\mathbb{R})$. Physically, \eqref{symp-matrix} is the statement that the new coordinates $\tilde{\xi}(\xi)$ again satisfy the fundamental brackets \eqref{fundamental-PB}: canonical transformations are exactly the changes of phase-space variables that leave the fundamental brackets, and hence \emph{all} Poisson brackets, intact. (Sanity check: the point transformations $\tilde{q}(q)$ with momenta co-transformed by \eqref{cotangent-transform} are canonical, as the reader should verify from \eqref{symp-matrix} with block-diagonal-plus-nilpotent $\Lambda$; but the canonical group is vastly larger, freely mixing positions and momenta, e.g. $(q,p)\mapsto(p,-q)$.)

Four propositions organize the basic theory; the proofs are short enough to give in full.

\textbf{Proposition 1.} \emph{Symplectic transformations preserve all Poisson brackets:} $\cbr{f\circ\varphi,\, g\circ\varphi} = \cbr{f,g}\circ\varphi$.

\emph{Proof.} The bracket is built from $\Omega$ (via $\cbr{f,g} = \Omega(X_g,X_f)$, or in coordinates from $\tilde{\Omega}^{\alpha\beta}$), and pullback commutes with every ingredient: $\varphi^*$ of a contraction is the contraction of the pullbacks, and $\varphi^*\Omega = \Omega$. In coordinates it's \eqref{symp-matrix} sandwiched between two transformed gradients. $\blacksquare$

\textbf{Proposition 2.} \emph{Symplectic transformations map Hamiltonian flows to Hamiltonian flows: the $\varphi$-image of the flow of $H$ is the flow of $H\circ\varphi^{-1}$.}

\emph{Proof.} Push the field $X_H$ forward by $\varphi$ and test it in the defining equation: for the pushed field $\varphi_*X_H$,
\begin{flalign}
    i_{\varphi_* X_H}\,\Omega &= i_{\varphi_* X_H}\,\br{\varphi^{-1}}^*\!\!\;\varphi^*\,\Omega
    = \br{\varphi^{-1}}^*\br{i_{X_H}\,\varphi^*\Omega} \notag \\
    &= \br{\varphi^{-1}}^*\br{i_{X_H}\Omega} = -\br{\varphi^{-1}}^*\dd{H} = -\dd{\br{H\circ\varphi^{-1}}},
\end{flalign}
using \eqref{symptrans-def} and naturality of $i$ and $\dd{}$ under diffeomorphisms. So $\varphi_* X_H = X_{H\circ\varphi^{-1}}$. $\blacksquare$

This is the invariance property that makes canonical transformations \emph{useful}: transform to variables in which $H$ looks simple (action--angle variables being the apotheosis), solve there, transform back, with the guarantee that Hamiltonian structure survives every step.

\textbf{Proposition 3.} \emph{The symplectic transformations of $(\mathcal{M},\Omega)$ form a group under composition.}

\emph{Proof.} $(\varphi\circ\psi)^* = \psi^*\circ\varphi^*$ (contravariance of pullback), so if both factors preserve $\Omega$ so does the composite; the identity trivially qualifies; and applying $(\varphi^{-1})^*$ to \eqref{symptrans-def} gives $\Omega = (\varphi^{-1})^*\Omega$, so inverses qualify. $\blacksquare$

\textbf{Proposition 4.} \emph{Hamiltonian flows are one-parameter groups of symplectic transformations: for every observable $f$ and parameter $s$, $\br{\varphi^f_s}^*\,\Omega = \Omega$.}

This is the fundamental theorem of the section, the statement that dynamics itself is a canonical transformation, continuously unfolding, and Cartan's formula makes the proof three lines.

\emph{Proof.} It suffices to show the derivative in $s$ vanishes, i.e. $\Lie_{X_f}\Omega = 0$ (then $\br{\varphi^f_s}^*\Omega$ is $s$-independent and equals its value $\Omega$ at $s=0$). By Cartan \eqref{cartan},
\begin{flalign}
    \Lie_{X_f}\,\Omega = i_{X_f}\,\underbrace{\dd{\Omega}}_{=\,0} + \; \dd{\underbrace{\br{i_{X_f}\Omega}}_{=\,-\dd{f}}} = -\dd{}^2 f = 0, \label{flow-preserves}
\end{flalign}
using closedness of $\Omega$ in the first slot, the defining equation \eqref{hamvf} in the second, and $\dd{}^2 = 0$. $\blacksquare$

Look at the anatomy of \eqref{flow-preserves}: \emph{this} is where closedness of the symplectic form finally earns its keep, exactly as promised in \S\ref{sec:coordfree}. Nondegeneracy made Hamiltonian dynamics exist; closedness makes it symplectic. The converse of Proposition 4 is the local-Hamiltonian criterion \eqref{locham} again: a flow preserves $\Omega$ iff $\dd{(i_X\Omega)} = 0$ iff $X$ is locally Hamiltonian, tying the whole circle of ideas shut.

\subsection{Integral invariants}

Proposition 4 has quantitative teeth. If $\varphi^*\Omega = \Omega$ then also $\varphi^*(\Omega^k) = (\varphi^*\Omega)^k = \Omega^k$ for every wedge power $k = 1, \dots, n$, and hence for any $2k$-dimensional surface $S \subset \mathcal{M}$,
\begin{flalign}
    \int_{\varphi(S)} \Omega^k = \int_S \varphi^*\br{\Omega^k} = \int_S \Omega^k : \label{integral-invariants}
\end{flalign}
the integrals of all powers of $\Omega$ are \emph{integral invariants} of every canonical transformation, in particular of Hamiltonian time evolution. The $k = 1$ case has a vivid reading in canonical coordinates: $\int_S \Omega = \sum_a \int_S \dd{p_a}\wedge\dd{q^a}$ is the \emph{sum of the oriented areas of the shadows of $S$} on the $n$ canonical planes $(q^a, p_a)$, and this sum of shadow-areas is frozen under the dynamics however wildly $S$ itself is stretched and folded. That is the precise cash value of the slogan from \S\ref{sec:forms}: symplectic geometry is the geometry of area, and Hamiltonian dynamics is area-preserving in exactly this sense. For closed surfaces there is a companion \emph{relative} invariant: if $C = \partial S$ is a closed curve, Stokes gives $\oint_C \theta = \int_S \dd{\theta} = \int_S\Omega$ (on a phase space where $\Omega = \dd{\theta}$), so the loop integral of the action form $\oint p_a\,\dd{q^a}$ is likewise preserved, the classical quantity that Bohr--Sommerfeld quantization later declares to be $2\pi\hbar$ times an integer. The $k = n$ case of \eqref{integral-invariants} is a statement about volume, and it's important enough to be the next section.

\section{Liouville's Theorem}\label{sec:liouville}

\subsection{The symplectic volume and the theorem}

On a symplectic manifold of dimension $2n$, the top wedge power of $\Omega$ is a top form, hence a candidate volume element; nondegeneracy makes it an actual one (nowhere vanishing). Normalize:
\begin{empheq}[box=\mathbox]{align}
    v \equiv \frac{1}{n!}\;\underbrace{\Omega\wedge\Omega\wedge\cdots\wedge\Omega}_{n\ \text{factors}}, \label{liouville-volume}
\end{empheq}
the \emph{Liouville volume}. In canonical coordinates, expanding $(\dd{p_a}\wedge\dd{q^a})^{\wedge n}$: the cross terms with any repeated index die by $\dd{p_a}\wedge\dd{p_a} = 0$, the surviving terms are the $n!$ orderings of the distinct pairs, all equal after reshuffling (each pair $\dd{p_a}\wedge\dd{q^a}$ is a $2$-form, and $2$-forms commute past each other under $\wedge$), and the $1/n!$ eats the multiplicity:
\begin{flalign}
    v = \dd{p_1}\wedge\dd{q^1}\wedge\cdots\wedge\dd{p_n}\wedge\dd{q^n},
\end{flalign}
the ordinary coordinate volume $\dd[n]{q}\,\dd[n]{p}$ up to a conventional sign from reordering. So ``phase space volume'', the object of statistical mechanics, is not extra structure: it is manufactured from the symplectic form alone, and every symplectic manifold comes equipped with it.

\textbf{Theorem (Liouville).} \emph{Hamiltonian flows preserve the Liouville volume:}
\begin{flalign}
    (\varphi^H_t)^*\,v = v, \qquad \text{i.e.} \qquad \mathrm{vol}\!\br{\varphi^H_t(R)} = \mathrm{vol}(R) \quad \text{for every region $R$ of phase space.}
\end{flalign}

\emph{Proof.} $\br{\varphi_t^H}^* v = \frac{1}{n!}\br{\varphi_t^{H*}\Omega}^{\wedge n} = \frac{1}{n!}\,\Omega^{\wedge n} = v$ by Proposition 4. $\blacksquare$

That's the whole proof; it's the $k=n$ instance of \eqref{integral-invariants}. A blob of initial conditions evolves like a drop of incompressible fluid: it may deform grotesquely, filament, and fold (and for chaotic dynamics it does exactly that), but its volume is untouchable. Dissipation, in the phase-space sense of volumes contracting onto attractors, is structurally impossible for Hamiltonian dynamics; this single theorem is the reason equilibrium statistical mechanics (which needs an invariant measure to define ensembles on) is possible at all.

\subsection{The divergence-theorem proof}

Because Liouville's theorem is so load-bearing, let's prove it a second time, with pre-geometric tools, following Arnold. The rate of change of the volume of an evolving region is the flux of the velocity field through its boundary, i.e. the integral of the divergence (the transport/divergence theorem of vector calculus):
\begin{flalign}
    \dv{t}\,\mathrm{vol}\br{R_t} = \int_{R_t} \div X \; \dd[2n]{\xi}, \qquad \div X \equiv \partial_\alpha X^\alpha.
\end{flalign}
(If one wants this from scratch: $\mathrm{vol}(R_t) = \int_{R_0}\det\pdv*{\varphi_t(\xi)}{\xi}\,\dd[2n]{\xi}$, expand $\varphi_t(\xi) = \xi + tX(\xi) + O(t^2)$, and use the linearization $\det(\mathbbm{1} + tA) = 1 + t\tr A + O(t^2)$, which follows by expanding the determinant and noting every other term is at least quadratic in $t$; the trace of the Jacobian of $X$ is $\div X$.) So the theorem reduces to:

\textbf{Lemma.} Hamiltonian vector fields are divergence-free.

\emph{Proof.} From \eqref{hameqs-xi}, $X_H^\alpha = \tilde{\Omega}^{\alpha\beta}\partial_\beta H$ with constant $\tilde{\Omega}$, so
\begin{flalign}
    \div X_H = \partial_\alpha\br{\tilde{\Omega}^{\alpha\beta}\,\partial_\beta H} = \tilde{\Omega}^{\alpha\beta}\,\partial_\alpha\partial_\beta H = 0,
\end{flalign}
an antisymmetric matrix contracted with symmetric second partials. $\blacksquare$

Same cancellation as ever, symmetric-against-antisymmetric, now doing volume conservation. In the $q,p$ blocks the lemma reads $\pdv*{\dot{q}^a}{q^a} + \pdv*{\dot{p}_a}{p_a} = \pdv*{}{q^a}\pdv*{H}{p_a} - \pdv*{}{p_a}\pdv*{H}{q^a} = 0$: the phase-space velocity field is incompressible because its $q$-divergence and $p$-divergence are the same mixed derivative with opposite signs. (Contrast a damped oscillator, $\dot{q} = p/m$, $\dot{p} = -kq - \gamma p$: divergence $-\gamma < 0$, volumes shrink, the flow is not Hamiltonian for any $H$, which the geometry knew all along.)

\subsection{Poincar\'e recurrence}

Volume preservation plus bounded phase space has a startling consequence, worth including both for its own sake and as a specimen of how much follows from measure-level information alone.

\textbf{Theorem (Poincar\'e recurrence).} Let $\varphi$ be a volume-preserving invertible map of a bounded region $D$ to itself ($\varphi(D) = D$). Then for any neighbourhood $U$ (however small) of any point of $D$, there is a point of $U$ that returns to $U$ after finitely many applications of $\varphi$.

\emph{Proof.} Consider the images $U, \varphi(U), \varphi^2(U), \dots$. All have the same (positive) volume as $U$. If they were pairwise disjoint, $D$ would contain infinitely many disjoint sets of equal positive volume, contradicting $\mathrm{vol}(D) < \infty$. So some pair intersects: $\varphi^k(U)\cap\varphi^\ell(U)\neq\emptyset$ with $k > \ell \geq 0$. Apply $\varphi^{-\ell}$ (invertibility): $\varphi^{k-\ell}(U)\cap U \neq \emptyset$. Pick $y$ in that intersection: then $y \in U$, and $x \equiv \varphi^{-(k-\ell)}(y)$ lies in $U$ (since $y \in \varphi^{k-\ell}(U)$) with $\varphi^{k-\ell}(x) = y \in U$. So $x$ is a point of $U$ that returns to $U$ after $k - \ell$ steps. $\blacksquare$

Apply it to the time-$\tau$ map of a Hamiltonian flow on a bounded energy shell (bounded, e.g., for confining potentials): \emph{almost every state returns arbitrarily close to itself, infinitely often}. Now the famous paradox: gas in the left half of a box, partition removed, gas spreads; recurrence says the all-molecules-back-in-the-left-half configuration will be revisited to any accuracy. It will. The resolution is quantitative, not logical: the recurrence time for macroscopic $N$ is stupendous (estimates run to $10^{10^{N}}$-ish multiples of any human timescale), and the theorem is silent about \emph{when}. The moral for us is structural: Hamiltonian dynamics is, at the level of the invariant measure, \emph{reversible and recurrent}, and the arrow of time in thermodynamics must therefore be a statement about initial conditions and timescales, not about the mechanics. The mechanics has no arrow; we've just proven it.

\section{Statistical Distributions and the Liouville Equation}\label{sec:statistical}

\subsection{Distributions on phase space}

A single classical state is a point of $\Gamma$. But points are an idealization: finite-precision preparation, or deliberate statistical description, hands us instead a \emph{probability distribution} $\rho(\xi, t) \geq 0$ on phase space, normalized against the Liouville volume,
\begin{flalign}
    \int_\Gamma \rho\;\, v = 1
\end{flalign}
(the geometry of \S\ref{sec:liouville} is what makes this well-defined without choosing coordinates), with expectation values
\begin{flalign}
    \langle f \rangle_\rho = \int_\Gamma f\, \rho\;\, v. \label{expectation}
\end{flalign}
A point state is the limiting case $\rho \to \delta$-function; everything in this section degenerates correctly in that limit.

\subsection{The Liouville equation}

How does $\rho$ evolve? The probability ascribed to a co-moving blob must be constant (states neither appear nor disappear; they just move), and the blob's volume is constant by Liouville's theorem; therefore \emph{the density along a trajectory is constant}:
\begin{flalign}
    0 = \dv{t}\,\rho\br{\xi(t), t} = \pdv{\rho}{t} + \dot{\xi}^\alpha\,\partial_\alpha \rho = \pdv{\rho}{t} + \cbr{\rho, H},
\end{flalign}
substituting Hamilton's equations \eqref{hameqs-xi} and recognizing the bracket \eqref{PB}. Rearranged:
\begin{empheq}[box=\mathbox]{align}
    \pdv{\rho}{t} = -\cbr{\rho, H} = \cbr{H, \rho}. \label{liouville-eq}
\end{empheq}
This is the \emph{Liouville equation}, the master evolution law of classical statistical mechanics. Note the sign relative to observable evolution \eqref{PB-evolution}: an observable evolves as $\dot{f} = +\cbr{f, H}$, the distribution as $\dot{\rho} = -\cbr{\rho,H}$. The two evolutions are dual through the pairing \eqref{expectation}, and the opposite signs are exactly what keeps expectation values consistent whether we evolve the observable and freeze the state or vice versa,
\begin{flalign}
    \dv{t}\,\langle f\rangle = \int f\,\cbr{H,\rho}\, v = \int \cbr{f, H}\,\rho\; v = \langle \cbr{f,H}\rangle
\end{flalign}
(the middle step moves the bracket across the integral by parts, using $\div X_H = 0$ from \S\ref{sec:liouville} to kill the boundary/divergence terms). The reader who has seen the Schr\"odinger--Heisenberg duality of quantum mechanics should be experiencing d\'ej\`a vu, and the next paragraph makes it official. Stationary distributions, $\pdv*{\rho}{t} = 0$, are those with $\cbr{\rho, H} = 0$, in particular any $\rho = F(H)$ (functions of conserved quantities); the microcanonical and canonical ensembles of statistical mechanics are exactly such choices, sitting at the end of this chain of reasoning as the invariant measures Liouville's theorem promised us.

\subsection{The bridge to quantum mechanics}

Assemble the classical structures in one column and their quantum counterparts in the other. Observables: functions on phase space $\longleftrightarrow$ Hermitian operators on Hilbert space. Bracket: $\cbr{f,g} \longleftrightarrow \frac{1}{i\hbar}[\hat{f},\hat{g}]$ (Dirac's correspondence, with the $i\hbar$ mandatory: see the box in \S\ref{sec:poisson}). Fundamental brackets \eqref{fundamental-PB}: $\cbr{q^a,p_b} = \delta^a_b \longleftrightarrow [\hat{q}^a,\hat{p}_b] = i\hbar\,\delta^a_b$. Generators and flows (\S\ref{sec:hamvf}): every observable generates a one-parameter group of canonical transformations $\longleftrightarrow$ every Hermitian operator generates a one-parameter unitary group $e^{-is\hat{f}/\hbar}$. Constants trivial in the generator (\S\ref{sec:hamvf}) $\longleftrightarrow$ global phase invisible. Observable evolution $\dot f = \cbr{f,H}$ $\longleftrightarrow$ Heisenberg's $\dv*{\hat{A}_H}{t} = \frac{1}{i\hbar}[\hat{A}_H, \hat{H}]$. State evolution \eqref{liouville-eq} $\longleftrightarrow$ von Neumann's $\pdv*{\hat\rho}{t} = \frac{1}{i\hbar}[\hat H, \hat\rho]$, sign flip included. Liouville's theorem, evolution preserves the measure $\longleftrightarrow$ unitarity, evolution preserves the inner product. The dictionary is too systematic to be coincidence, and it isn't one: quantization, in every one of its formulations (canonical, path-integral, deformation, geometric), is an attempt to promote precisely the symplectic structures of these notes to operator or amplitude form, and the classical column is not scaffolding to be discarded but the $\hbar \to 0$ skeleton the quantum theory is built on. Which is where these notes hand off to their companions: the path integral is what the tautological form's $\int p\,\dd q - H\,\dd t$ becomes when Dirac's ``$e^{iS/\hbar}$ corresponds to the propagator'' is taken with full seriousness, and the reader is warmly pointed there.

\end{document}